
---
title: "PCA"
author: |
  Jyotishka Datta\\
  Department of Statistics, Virginia Tech
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  beamer_presentation:
    theme: "Madrid"
    slide_level: 2
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

## Source

- A lot of R code in this demo is adapted from the excellent book *An Introduction to Statistical Learning with Applications in R* by James, Witten, Hastie, and Tibshirani. The book website provides R code, labs, and other resources.
- The Crime and Iris data examples were influenced by materials from Dr. Sumanta Basu. PCR code & examples were provided by Dr. Bobby Gramacy.

## Recap (PCA)

Let \(\mathbf{X}\) be an \(n \times p\) data matrix with columns centered (column means subtracted).

**First principal component (PC1)** is the direction \(\boldsymbol{\beta}_1 = (\beta_{11}, \dots, \beta_{p1})^\top\) that maximizes the variance of the projected data, subject to unit length:
\[
\max_{\boldsymbol{\beta}_1} \; \frac{1}{n} \sum_{i=1}^n (\mathbf{x}_i^\top \boldsymbol{\beta}_1)^2
\quad \text{s.t.} \quad \lVert \boldsymbol{\beta}_1 \rVert_2 = 1.
\]
The scores are \(y_{i1} = \mathbf{x}_i^\top\boldsymbol{\beta}_1\).

## Second PC

The second principal component \(\boldsymbol{\beta}_2\) solves
\[
\max_{\boldsymbol{\beta}_2} \; \frac{1}{n} \sum_{i=1}^n (\mathbf{x}_i^\top \boldsymbol{\beta}_2)^2
\quad \text{s.t.} \quad \lVert \boldsymbol{\beta}_2 \rVert_2 = 1,\; \boldsymbol{\beta}_1^\top\boldsymbol{\beta}_2 = 0.
\]
Scores \(y_{i1}\) and \(y_{i2}\) are uncorrelated by construction.

## Optimization for PCA

Two equivalent computational routes for PCA:

- **SVD of \(\mathbf{X}\)**: \(\mathbf{X} = \mathbf{U}\,\boldsymbol{\Sigma}\,\mathbf{V}^\top\). The columns of \(\mathbf{V}\) are PC loading vectors \(\boldsymbol{\beta}_j\); singular values give standard deviations of PCs.
- **Eigen-decomposition of covariance**: On centered data, \(\mathbf{X}^\top\mathbf{X}\) eigenvectors are loadings, eigenvalues are variances of PCs.

## Using R

Two common functions:

- `prcomp()` uses an SVD on the (centered/scaled) data matrix (numerically stable).
- `princomp()` uses an eigen-decomposition of the covariance (or correlation) matrix.

## US Arrests data

**Description.** Statistics in arrests per 100,000 residents for Assault, Murder, and Rape in each of the 50 US states in 1973. Also given is percent of the population in urban areas (`UrbanPop`).

```{r usarrests-setup}
states <- row.names(USArrests)
states
names(USArrests)
apply(USArrests, 2, mean)
apply(USArrests, 2, var)
```

> If we fail to scale variables before PCA, components can be dominated by variables with the largest variance (here, `Assault`).

## PCA on USArrests

```{r usarrests-pca}
pr.out <- prcomp(USArrests, scale. = TRUE)
names(pr.out)
pr.out$center
pr.out$scale
```

## Loadings and scores

```{r usarrests-loadings}
pr.out$rotation  # loadings (each column is a PC loading vector)
dim(pr.out$x)    # 50 observations x 4 PCs
```

## Biplot

A biplot visualizes both observations (scores) and variables (loadings).

```{r usarrests-biplot, fig.height=5.5}
biplot(pr.out, scale = 0, cex = c(0.6, 0.75), xlim = c(-4, 4))
```

## Sign indeterminacy

Principal components are unique **up to sign**. Flipping the sign of a loading vector and its scores leaves the geometry unchanged.

```{r usarrests-biplot-sign}
pr.tmp <- pr.out
pr.tmp$rotation <- -pr.tmp$rotation
pr.tmp$x <- -pr.tmp$x
biplot(pr.tmp, scale = 0, cex = c(0.5, 1), xlim = c(-4, 4))
```

## Importance of scaling

```{r usarrests-noscale}
biplot(prcomp(USArrests, scale. = FALSE), scale = 0, cex = c(0.5, 1))
```

## Standard deviations and PVE

```{r usarrests-sdev-pve}
pr.out$sdev
pr.var <- pr.out$sdev^2
pr.var
pve <- pr.var / sum(pr.var)
pve
```

## Proportion of Variance Explained (PVE)

```{r usarrests-pve-plot}
par(mfrow = c(1, 2))
barplot(pve, main = "Prop. Variance Explained", ylab = "PVE", ylim = c(0, 1))
plot(cumsum(pve), type = "b", pch = 19, xlab = "Number of PCs", ylab = "Cumulative PVE",
     ylim = c(0, 1))
par(mfrow = c(1, 1))
```

## More examples (`princomp`)

`prcomp` (SVD on data) and `princomp` (eigen on covariance/correlation) usually agree closely; some prefer `prcomp` for numerical stability.

## PCA for City Crime data

The city crime dataset contains crime rates per 100,000 people for 72 large US cities in 1994. Variables: Murder, Rape, Robbery, Assault, Burglary, Larceny, Motor Vehicle Theft.

```{r crime-read}
# Place a file named "city_crime.txt" in the project directory.
# The file should have 7 numeric columns in the order listed above.
crime <- NULL
if (file.exists("city_crime.txt")) {
  crime <- tryCatch({
    read.table("city_crime.txt", header = TRUE)
  }, error = function(e) tryCatch({
    read.table("city_crime.txt", header = FALSE)
  }, error = function(e2) NULL))
}
if (is.null(crime)) {
  message("city_crime.txt not found or unreadable; subsequent crime PCA plots are skipped.")
}
```

## Scatterplot matrix (crime)

```{r crime-pairs, eval=!is.null(crime)}
pairs(crime)
```

## PCA on crime data (`princomp`)

```{r crime-princomp, eval=!is.null(crime)}
pca.crime <- princomp(scale(crime), cor = FALSE)
summary(pca.crime)
loadings(pca.crime)
```

## PVE and scree (crime)

```{r crime-pve-scree, eval=!is.null(crime)}
PoV <- pca.crime$sdev^2 / sum(pca.crime$sdev^2)
barplot(PoV, main = "Prop Var Exp")
screeplot(pca.crime, type = "line")
```

## Scores and biplot (crime)

```{r crime-scores-biplot, eval=!is.null(crime)}
pcs.crime <- predict(pca.crime)
plot(pcs.crime[, 1:2], type = "n", xlab = '1st PC', ylab = '2nd PC')
text(pcs.crime[, 1:2], row.names(crime))
biplot(pca.crime, cex = c(0.5, 1))
```

## Questions

1. Which cities appear most safe? Which are most crime-prone?
2. How might we interpret high/low values on the second PC?

## Subtle difference: scaling in `prcomp` vs `princomp`

```{r crime-scale-diff, eval=!is.null(crime)}
pc.cr  <- prcomp(crime, scale = TRUE, center = TRUE)
pc.cr1 <- princomp(crime, cor = TRUE, scores = TRUE)
pc.cr$scale
pc.cr1$scale
```

## Iris data

We will check if PCA can still detect species after adding noise variables.

```{r iris-setup}
data(iris)
str(iris)
# Matrix of original 4 features
X_iris <- data.matrix(iris[, 1:4])
# Add 6 columns of i.i.d. N(0,1) noise as extra dimensions
set.seed(2023)
noise <- matrix(rnorm(nrow(X_iris) * 6), ncol = 6)
# Combine so that noise appears as first 6 columns
dat.iris <- cbind(noise, X_iris)
colnames(dat.iris) <- c(paste0('V', 1:6), colnames(X_iris))
# Standardize before PCA on correlation structure
dat.iris <- scale(dat.iris, center = TRUE, scale = TRUE)
pca.iris <- princomp(dat.iris, cor = TRUE)
```

## Iris PCA: summary and scree

```{r iris-summary}
summary(pca.iris)
screeplot(pca.iris, type = "line")
```

## Iris PC scores and colors

```{r iris-scores-plot}
pcs.iris <- predict(pca.iris)
col.code <- as.character(iris$Species)
col.code[col.code == "setosa"]     <- 'red'
col.code[col.code == "versicolor"] <- 'blue'
col.code[col.code == "virginica"]  <- 'green'
plot(pcs.iris[, 1:2], type = 'p', pch = 15, col = col.code,
     xlab = "PC1", ylab = "PC2")
legend("topright", col = c('red','blue','green'),
       legend = c('setosa','versicolor','virginica'), pch = 15)
```

## Why PCs can help

Compare to plotting just two (arbitrary) variables in the original space.

```{r iris-compare}
plot(dat.iris[, 1:2], type='p', pch=15, col=col.code, xlab="X1", ylab="X2")
legend("topright", col=c('red','blue','green'), legend=c('setosa','versicolor','virginica'), pch=15)
# Re-plot PC space to emphasize separation
plot(pcs.iris[, 1:2], type='p', pch=15, col=col.code, xlab="PC1", ylab="PC2")
legend("bottomleft", col=c('red','blue','green'), legend=c('setosa','versicolor','virginica'), pch=15)
```

## Principal Components Regression (PCR)

PCR uses the leading PC score columns \(z_1,\dots,z_m\) (with \(m \ll p\)) as predictors in a regression for response \(y\). With orthogonal PCs, the OLS fit decomposes into a sum of univariate regressions in the PC scores.

## PCR: backing out coefficients

Each PC is a linear combination of the original (scaled) variables, so PCR coefficients in the original feature space can be recovered from the PC loadings and the regression on PC scores.

## Toy PCR example (olive oil)

```{r pcr-oliveoil}
# Ensure pls is available
if (!requireNamespace("pls", quietly = TRUE)) {
  install.packages("pls", repos = "https://cloud.r-project.org")
}
library(pls)
data(oliveoil)
# Response: first sensory score; Predictors: chemical measurements
set.seed(1)
y <- as.numeric(oliveoil$sensory[, 1])
X <- as.matrix(oliveoil$chemical)
# Standardize predictors
for (j in 1:ncol(X)) X[, j] <- X[, j] / sd(X[, j])
# PCA via eigen decomposition of covariance
S  <- cov(X)
eS <- eigen(S, symmetric = TRUE)
Z  <- X %*% eS$vectors
colnames(Z) <- paste0("PC", 1:ncol(Z))
# Fraction of variance explained
cumsum(eS$values) / sum(eS$values)
# Use first three PCs (plus intercept) for regression
Zp <- cbind(1, Z[, 1:3, drop = FALSE])
theta.pcr <- drop(solve(t(Zp) %*% Zp) %*% t(Zp) %*% y)
# Map back to original coefficients (including intercept)
beta.pcr <- c(theta.pcr[1], eS$vectors[, 1:3, drop = FALSE] %*% theta.pcr[-1])
names(beta.pcr) <- c("Intercept", colnames(oliveoil$chemical))
beta.pcr
```

## PCR via `pls::pcr()` and CV

```{r pcr-pls}
fit.pcr <- pcr(y ~ X, ncomp = 3)
rbind(automated = coef(fit.pcr, intercept = TRUE), byhand = as.numeric(beta.pcr))
cv.pcr <- pcr(y ~ X, validation = "LOO")
RMSEP(cv.pcr)
plot(RMSEP(cv.pcr), main = "")
```

---

### Reproducibility notes

- Rendering this file to PDF requires a LaTeX installation. If you use R, you can install TinyTeX via:
  
  ```r
  if (!requireNamespace("tinytex", quietly = TRUE)) install.packages("tinytex"); tinytex::install_tinytex()
  ```
- To include the city crime example, add `city_crime.txt` to the working directory.
- Tested with R \u2265 4.0 and packages in CRAN.
